Covariate analysis of HBCC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN + SVA normlaisation)

### 2018-07-25 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Modifications in HBCC DLPFC SVA markdown to account for individual ID changes in metadata

knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
                                 parentId = "syn7501880", #syn9872607",
                                 entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn7501880'; #'syn9872607';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (DLPFC)';

>>>>

knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
                                 parentId = "syn9872607", #"syn7501880", 
                                 entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn9872607'; #'syn7501880'; #
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (DLPFC)';

### 2018-07-25 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
Modifications in HBCC DLPFC SVA markdown to account for individual ID changes in metadata

knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd",
                                 parentId ="syn7501880",
                                 entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - SVA')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn7501880';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN and SVA normalisation (DLPFC)';

@@

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download reprocessed counts (DLPFC)
COUNT_DLPFC_ID = 'syn8465067';
ALL_USED_IDs = COUNT_DLPFC_ID
COUNT = downloadFile(COUNT_DLPFC_ID) %>%
  as.data.frame()

GENE.LEN = dplyr::select(COUNT, GeneID, Length) %>%
  dplyr::rename(ensembl_gene_id = GeneID) %>%
  unique()
rownames(GENE.LEN) = GENE.LEN$ensembl_gene_id

rownames(COUNT) = COUNT$GeneID
ind = setdiff(colnames(COUNT), c('GeneID', 'Chr', 'Start', 'End', 'Strand', 'Length'))
COUNT = data.matrix(COUNT[,ind])

# Get ancestry vector calculated using gemtools
# ANCESTRY_ID = 'syn2511399'
# ALL_USED_IDs[length(ALL_USED_IDs)+1] = ANCESTRY_ID
# ANCESTRY = read.table(synGet(ANCESTRY_ID)@filePath, fill = T, header = T, sep = '\t') %>%
#   plyr::rename(c('DNA_report..Genotyping.Sample_ID' = 'GenotypingSampleID'))

# Get RNA qc metrics from synapse
METADATA_ALG_REP_DLPFC_ID = 'syn2279446'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ALG_REP_DLPFC_ID
METADATA_ALG_REP_DLPFC = read.table(synGet(METADATA_ALG_REP_DLPFC_ID)@filePath, 
                                    fill = T, header = T, sep = ',') %>%
  dplyr::select(Sample.RNA.ID, Ribozero.Batch, Library.Batch, Flowcell.Batch) %>%
  dplyr::rename(SampleID = Sample.RNA.ID,
                LibraryBatch = Library.Batch, 
                RibozeroBatch = Ribozero.Batch, 
                FlowcellBatch = Flowcell.Batch) %>%
  unique()

# Get RNA qc metrics from synapse
METADATA_QC_DLPFC_ID = 'syn8539839'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_DLPFC_ID
METADATA_QC = read.table(synGet(METADATA_QC_DLPFC_ID)@filePath, fill = T, header = T, sep = ',', row.names = 1) %>%
  dplyr::select(Sample, Total_Purity_Filtered_Reads_Sequenced, X5prime_Norm, Mean_Per_Base_Cov.,  
                Mean_CV, Fragment_Length_StdDev,  Mapped_Pairs, End_1_Mismatch_Rate, End_2_Mismatch_Rate,
                End_1_Mapping_Rate, End_2_Mapping_Rate, Mapped, Intragenic_Rate, Exonic_Rate, Intergenic_Rate,
                Intronic_Rate, rRNA_rate, Genes_Detected, Transcripts_Detected, Expression_Profiling_Efficiency,
                UNIQUE_READS._Uniquely_mapped_reads_percent) %>%
  dplyr::rename(SampleID = Sample, 
                IntragenicRate = Intragenic_Rate, 
                ExonicRate = Exonic_Rate, 
                IntronicRate = Intronic_Rate, 
                IntergenicRate = Intergenic_Rate, 
                GenesDetected = Genes_Detected, 
                TranscriptsDetected = Transcripts_Detected, 
                ExpProfEfficiency = Expression_Profiling_Efficiency, 
                rRNARate = rRNA_rate, 
                TotalReads = Total_Purity_Filtered_Reads_Sequenced, 
                PercentMapped = UNIQUE_READS._Uniquely_mapped_reads_percent)

# Get merged metadata
METADATA_ID = 'syn2299154'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
METADATA = read.csv(synGet(METADATA_ID)@filePath, fill = T, header = T, check.names = F)

## Metadata with specific information for RNASeq processing
metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", "Dx",

                          "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?")

# Filter metadata 
ind = colnames(METADATA) %in% metadataVarOfInterest
METADATA = METADATA[,ind]
colnames(METADATA) = gsub('[^[:alnum:]]','_',colnames(METADATA))

METADATA = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
                DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, 
                DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
                RIN = DLPFC_RNA_isolation__RIN, 
                ReportExclude = DLPFC_RNA_report__Exclude_) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'DLPFC') %>%
  list(.,METADATA_QC, METADATA_ALG_REP_DLPFC) %>%
  join_all %>% droplevels() %>%
  unique %>% as.data.frame()
METADATA$SampleID = as.character(METADATA$SampleID)

# Fix missing values
levels(METADATA$LibraryBatch)[1] = 'NotAvailable'
levels(METADATA$FlowcellBatch)[1] = 'NotAvailable'
levels(METADATA$RibozeroBatch)[1] = 'NotAvailable'
```

### Data preprocessing
```{r preprocess.data}
# Remove samples marked as exclude and samples with no Gender, Ethnicity, PMI, and RIN
writeLines('Following counts are missing any metadata')
writeLines(paste(setdiff(colnames(COUNT), METADATA$SampleID), collapse = ', '))
METADATA <- METADATA %>% dplyr::filter(SampleID %in% colnames(COUNT))

writeLines('Following samples are marked exclude')
writeLines(paste(METADATA$SampleID[METADATA$IsolationExclude == 1 | METADATA$ReportExclude == 1], collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(IsolationExclude == 0, ReportExclude == 0)

@@

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r detect.outliers, fig.height=5, fig.width=5, results='asis', cache = FALSE}
indToRemove = c("CMC_HBCC_RNA_PFC_3402")

@@

### Sample clustering
```{r decompse.normalise.data, fig.height=6, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
PC <- prcomp(tmp, scale.=T, center = T)

@@

# Compute actual variance of all principal components
tmp = svd(RESIDUAL.GENE_EXPRESSION)
actual.var = tmp$d^2/sum(tmp$d^2)

# Compute permuted variance of all principal components
permuted.var = foreach(i = 1:20, .combine = rbind, .packages = c('dplyr')) %dopar% {
  tmp.residual = t(apply(RESIDUAL.GENE_EXPRESSION, 1, sample, replace = FALSE))
  tmp = svd(tmp.residual)
  permuted.var = tmp$d^2/sum(tmp$d^2)
}
permuted.var = apply(permuted.var, 2, mean)
NUM.SV = sum(actual.var >= permuted.var[1])

# Plot variance components
var.comp = data.frame(component = 1:length(actual.var), 
                      residual = actual.var, 
                      permuted.residual = permuted.var) %>%
  tidyr::gather(data, value, -component)
p = ggplot(var.comp %>% filter(component <= round(NUM.SV*1.2)), aes(x = component, y = value, color = data)) + geom_point()
p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')
p = p + geom_vline(xintercept=NUM.SV, linetype = 'dashed') + ylab('Fraction of Variance Explained')
p = p + theme(legend.position = c(0.8, 0.8), legend.title = element_blank())
p

# Estimate surrogate variables
SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$E, MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
rownames(SURR.VAR) = rownames(MODEL1)

# Estimate voom weights for dispersion control
VOOM.SVA.WEIGHTS = voom(NEW.COUNTS, 
                        design=cbind(MODEL1, SURR.VAR), plot=F)
VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E

# Fit linear model using new weights and new design
VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)
```

### Residual calculation
Calculate weighted residuals and add back "Dx.Tissue" to the residuals
```{r varsToAddBack}
# Residuals after normalisation
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
                                                  CQN.GENE_EXPRESSION$E)

# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DESIGN), value = T)
RESIDUAL.SVA.GENE_EXPRESSION = RESIDUAL.SVA.GENE_EXPRESSION + 
  VOOM.SVA.FIT$coefficients[,varsToAddIn] %*% t(VOOM.SVA.FIT$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.SVA.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
Clustering residual data
```{r decompse.normalise.data2, fig.height=6, fig.width=6, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.SVA.GENE_EXPRESSION, scale.=T, center = T)

>>>>

knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd",
                                 parentId = "syn7501880", #"syn9872607",
                                 entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - SVA')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = "syn7501880", #'syn9872607';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN and SVA normalisation (DLPFC)';

@@

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data preprocessing
```{r preprocess.data}
# Remove samples marked as exclude and samples with no Gender, Ethnicity, PMI, and RIN
writeLines('Following counts are missing any metadata')
writeLines(paste(setdiff(colnames(COUNT), METADATA$SampleID), collapse = ', '))
METADATA <- METADATA %>% dplyr::filter(SampleID %in% colnames(COUNT))

writeLines('Following samples are marked exclude')
writeLines(paste(METADATA$SampleID[METADATA$label == 'exclude'], collapse = ', '))
METADATA <- METADATA  %>% dplyr::filter(label == 'accept')

@@

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r detect.outliers, fig.height=8, fig.width=8, results='asis', cache = FALSE}
indToRemove = c("CMC_HBCC_RNA_PFC_3402")

@@

### Sample clustering
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
PC <- prcomp(tmp, scale.=T, center = T)

@@

# Compute actual variance of all principal components
expr = RESIDUAL.GENE_EXPRESSION; expr[is.na(expr)] = 0
tmp = svd(expr)
actual.var = tmp$d^2/sum(tmp$d^2)

# Compute permuted variance of all principal components
permuted.var = foreach(i = 1:20, .combine = rbind, .packages = c('dplyr')) %dopar% {
  tmp.residual = t(apply(expr, 1, sample, replace = FALSE))
  tmp = svd(tmp.residual)
  permuted.var = tmp$d^2/sum(tmp$d^2)
}
permuted.var = apply(permuted.var, 2, mean)
NUM.SV = sum(actual.var >= permuted.var[1])

# Plot variance components
var.comp = data.frame(component = 1:length(actual.var), 
                      residual = actual.var, 
                      permuted.residual = permuted.var) %>%
  tidyr::gather(data, value, -component)
p = ggplot(var.comp %>% 
             dplyr::filter(component <= round(NUM.SV*1.2)), 
           aes(x = component, y = value, color = data)) + geom_point()
p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')
p = p + geom_vline(xintercept=NUM.SV, linetype = 'dashed') + ylab('Fraction of Variance Explained')
p = p + theme(legend.position = c(0.8, 0.8), legend.title = element_blank())
p

# Estimate surrogate variables
expr = CQN.GENE_EXPRESSION$E
expr[is.na(expr)] = 0
SURR.VAR = irwsva.build(expr, MODEL1, MODEL0, 
                        n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
rownames(SURR.VAR) = rownames(MODEL1)

# Estimate voom weights for dispersion control
VOOM.SVA.WEIGHTS = voom(cnts, 
                        design=cbind(MODEL1, SURR.VAR), plot=F)
VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E

# Fit linear model using new weights and new design
VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)

# Residuals after normalisation
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
                                                  CQN.GENE_EXPRESSION$E)
```

### Residual calculation
Calculate weighted residuals and add back "Dx.Tissue" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DESIGN), value = T)
RESIDUAL.SVA.GENE_EXPRESSION = RESIDUAL.SVA.GENE_EXPRESSION + 
  VOOM.SVA.FIT$coefficients[,varsToAddIn] %*% t(VOOM.SVA.FIT$design[,varsToAddIn])
```
Clustering residual data
```{r decompse.normalise.data2, fig.height=6, fig.width=6, results='asis'}
# Find principal components of expression to plot
expr = RESIDUAL.SVA.GENE_EXPRESSION; expr[is.na(expr)] = 0
PC <- prcomp(expr, scale.=T, center = T)
